Predictive Models to Support Quoting of Fixed Fee Consulting Projects

### 2016-12-14 finalwriteup/Consulting Profitability Paper/Consulting_Profitability_Paper.tex
finished lit review paper

Financial risk most commonly falls upon a consulting company when they
must offer a fixed fee for a project. In competitive climates, clients
request fixed price quotes for project completion from several
consultants before a project is awarded. Hence, initial quoted fees must
be competitive but also accurately cover consultant salaries and
business costs.

How do consultants make this accurate yet competitive fee estimate? This
varies from industry to industry, but typically a consulting manager has
experience in the type of project they are quoting and, after reviewing
the project details, can use a combination of intuition, comparison to
past projects, and rules of thumb. Unfortunately, some projects priced
using this method can blow out of budget, causing significant losses to
the consultant.

A review of 6 IT project surveys between 1984 and 1994 by Moløkken \&
Jørgensen (2003) revealed that 60-80\% of IT projects encountered effort
and/or schedule overruns, where the average overrun was by 30 to 40\%. A
study on large-scale infrastructure projects over the past seventy years
revealed that cost forecasts consistently underestimated the cost of
rail projects by an average of 44.7\%, the cost of bridge and tunnel
projects by 33.8\% and the cost of road projects by 20.4\% (Flyvbjerg
2007).

@@

This work aims to provide a manager with this more realistic view by
predicting whether the new project will be profitable or not, based on a
suite of previous project data from the same company. The manager could
then take multiple actions including not tendering for a project
predicted to be loss-making, providing a high fixed price, or
restructuring the contract so that it is not 100\% fixed price.

@@

This thesis focuses on a single case study consulting company in the
construction industry. twelve years of passively collected data that
described each project in terms of their clients, invoice history,
employee hours and technical details. labour is the chief cost and the
long traditional history of this industry mandates fixed price projects
as the norm. 20\% of projects are loss making. Fees calculated via
detailed analysis A Customer Relationship Management software package
(CRM) is currently employed to collect and store project data. A CRM is
a popular type of software used by businesses to record client and
project details as well as employee time sheet records. In the case
study business, the CRM is available to all employees over the company
intranet, and each employee completes daily time sheets allotting their
hours to certain projects. Additionally, technical information is
recorded against each project

\begin{itemize}
\tightlist
\item
  Employee time sheet hours with dates
\item
  Other project costs (taxis, printing)
\item
  Client information/characteristics
\item
  Client identification code
\item
  Invoiced amounts for each project and dates
\item
  Employee costs
\item
  Employee charge out rates
\item
  Project description
\end{itemize}

2364 past projects are available. missing data due to how the data
collection changed over time and some fields were not mandatory.

To assess whether the model would improve the case study company's
profitability, a simple managerial decision framework was applied: if
the algorithm predicted the project to be loss making, the project was
considered `rejected' and all profits or \emph{losses} of rejected
projects were discarded. The remaining projects' profits and losses were
summed to give a revised overall profit over the last 12 years of
projects. This was compared to the summed profits and losses from all
projects completed over the last 12 years. The best algorithm rejected
enough loss-making projects to improve overall profits by 9\%.

\begin{itemize}
\tightlist
\item
  aim \textbf{General Aim} Use statistical techniques to model the
  profitability of projects for consulting businesses using their
  internal CRM data. Research will focus on a case study Engineering
  consulting company that offers their expert advice (in the currency of
  time) to business clients. The project outcomes are intended to assist
  the business in predicting project profitability before engagement.
  Several statistical and machine learning techniques were tested,
  compared and refined.
\end{itemize}

\textbf{Hypothesis 1} A statistical or machine learning model based on
historical project data can predict the profitability of a new project
with greater accuracy than a baseline predictor. A baseline predictor
for this project is one that predicts the average of a numeric response
variable for all cases or, if the response variable is categorical,
randomly assigned categories for each case, where the proportion of
assigned categories matches the true categorical proportions.

\textbf{Hypothesis 2} The predictive model built from Hypothesis 1 can
be shown to have a positive impact on the overall profit earned by the
case study business. The overall profit is represented by the following
equation:

@@

Summary of Key Related Research This section should include a brief
summary of key related research. Emphasis should be on demonstrating the
foundation for the current investigation. Specifically, the goal is to
clearly delineate a gap or missing link that the current research fills.
Authors should avoid presenting a litany of past research and should
focus on prior work necessary to demonstrate the existence of the
research gap addressed in the manuscript.

@@

The bulk of research to date has been performed with project data in
either the construction industry or software development. Research in
the construction industry has primarily focused on predicting the final
building cost of construction. Over the past few decades, significant
research has been dedicated to creating predictive models that present
the `outside view' of a project. This was done by statistically
comparing a new project to a collection of similar projects and their
characteristics. It has been shown that these models improve cost
estimation accuracy, however industry has not adopted them.

\subparagraph{2.1.1 Construction Industry}\label{construction-industry}

Elfaki, Alatawi, and Abushandi (2014) completed a review of cost
estimation research from 2004 to 2014. They found that artificial Neural
Networks and SVM's were the most common machine learning techniques.
Neural Networks and SVM's received significant attention in the 1990's
for their ability to accurately predict construction costs with limited
detailed information (Shin 2015; Kim, An, and Kang 2004). Shin (2015)
pioneered the application of Boosted Trees to cost estimation in
construction projects. Often, Linear Regression was the only model
assessed, without comparison to other methods such as Neural Networks,
which first started appearing in literature in the 1990's (Kim, An, and
Kang 2004) Some studies showed that Neural Networks outperform
Regression, however other studies established they are approximately
equal (Kim, An, and Kang 2004; Attalla and Hegazy 2003).

Variety of variables used: (Chan and Park 2005)(Akintoye and Fitzgerald
2000; Trost and Oberlender 2003; Pinto and Slevin 1988) the studies all
generally reported their mean absolute error or a similar metric from
the tested models, and results were generally positive. Furthermore, the
existing studies collected data from many businesses for a single study

\subparagraph{2.1.2 Software Industry}\label{software-industry}

In the software industry, the main component of cost is \emph{effort} as
opposed to the cost of building materials in the construction industry.

Similar to the construction industry, expert judgment or detailed
analysis, is the most widely practiced method for effort estimation
(Shepperd, Schofield, and Kitchenham 1996; Moløkken and Jørgensen 2003)

Multiple studies have shown that Neural Networks definitively outperform
regression models in effort estimation, although Regression is the most
popular method (Finnie, Wittig, and Desharnais 1997; Pai, McFall, and
Subramanian 2013; Matson and Mellichamp 1993).

Interestingly, several studies have found that even if 15 or so
variables are included, often only one variable contributes
significantly to the model's accuracy: size (or funciton point)
(Shepperd, Schofield, and Kitchenham 1996; Finnie, Wittig, and
Desharnais 1997; Pai, McFall, and Subramanian 2013)

A study by Mendes and Kitchenham (2004) demonstrated using 67 web
projects that cross-company models were significantly less accurate than
a within company model (Shepperd, Schofield, and Kitchenham 1996) Again,
Neural Networks were criticised for their inability to explain their
results (Finnie, Wittig, and Desharnais 1997). The software industry is
unique in that agile methods of delivery are changing the contractual
approach of consulting (Badenfelt 2011). Software development is a
relatively new field in comparison to construction, where the tradition
of fixed fee structures is difficult to reinvent.

\paragraph{2.2 Methods used in other business
applications}\label{methods-used-in-other-business-applications}

The review of methods applied to the construction and IT industry
highlighted the use of Linear Regression, Neural Networks, SVM's and in
one case Boosted Trees, however research on other business problems
utilised a wider range of methods. These included Naive Bayes, Random
Forests, and machine learned Bayesian Networks.

Although SVM's performed the best, as discussed, they are often complex
and slow, requiring a great deal of memory (Kumar and Ravi 2007). Kumar
and Ravi (2007) performed a detailed review of statistical and machine
learning techniques that were applied over 37 years in the context of
bankruptcy prediction in banks.

For example in financial credit scoring, a study by Brown and Mues
(2012) found that Random Forests and Boosted Trees consistently
outperformed Neural Networks in classification. Kumar and Ravi's (2007)
review also assessed ensemble techniques, which refers to combinations
of two completely different algorithms, and found they can often
outperform individual methods.

An exception is Saradhi \& Palshikar's (2011) study on employee churn,
where `churn' refers to the number of individuals moving out of a group
within a certain time. Naive Bayes, Random forests (an ensemble decision
tree method) and SVM's were built. determining the value of each
employee in terms of the importance of the projects they were on and
their monthly chargeability. This allowed them to rank employees
identified as `high risk of churn' by value and provided a clear ranking
for manager's to act upon. This extension of the study provided a
comprehensive framework for how business managers could adopt their
findings to improve business operations. It was a valuable addition that
is absent from most cost and effort estimation research.

@@

In comparison to previous studies, this project advances the body of
work on cost estimation in three ways: the data set is larger than other
studies and sourced from a single company's CRM database, ensemble tree
methods and model blending was tested, and the prediction results were
integrated into a decision framework for the business. In past research,
cost estimation data sets generally consisted of \textless{}300 of
projects from a range of a companies and even countries. This thesis
case study only had access to the company's internal data but this
contained over 2,000 past jobs Furthermore, the problem of cost
estimation is applied to a consulting company in the construction
industry, which differs from other studies in the construction industry
that estimate actual construction costs. Consulting in the construction
industry can be likened more to software development effort estimation
although the nature of the work differs considerably.

ensemble tree methods and blended methods are applied which have been
minimally tested on this problem. To date, Linear Regression, Neural
Networks, Case Based Reasoning (CBR), and Support Vector Machines
(SVM's) have been used to predict cost/effort even though ensemble tree
methods can perform as well as Neural Networks and generally outperform
Linear Regression (Caruana and Niculescu-Mizil 2006). Another benefit of
ensemble trees is the output types available, such as partial dependency
plots and variable importance plots that provide insight into the
model's calculations. This contrasts Neural Networks and SVM's, which
are `black-box' predictors.

Finally, this study propels the predictive model one step further than
other studies by analysing bottom-line profit improvements that could
result from the research. final profit increases are evaluated. This
kind of thought experiment is designed to present a clear case for
industry when evaluating if and how to adopt cost estimation models.

@@

\hypertarget{refs}{}
\hypertarget{ref-Akintoye2000}{}
Akintoye, Akintola, and Eamon Fitzgerald. 2000. ``A Survey of Current
Cost Estimating Practices in the Uk.'' Journal Article.
\emph{Construction Management and Economics} 18 (2): 161--72.
doi:\href{https://doi.org/10.1080/014461900370799}{10.1080/014461900370799}.

\hypertarget{ref-Attalla2003}{}
Attalla, Mohamed, and Tarek Hegazy. 2003. ``Predicting Cost Deviation in
Reconstruction Projects: Artificial Neural Networks Versus Regression.''
Journal Article. \emph{Journal of Construction Engineering and
Management} 129 (4): 405--11.

@@

\hypertarget{ref-Badenfelt2011}{}
Badenfelt, Ulrika. 2011. ``Fixing the Contract After the Contract Is
Fixed: A Study of Incomplete Contracts in It and Construction
Projects.'' Journal Article. \emph{International Journal of Project
Management} 29 (5): 568--76.
doi:\href{https://doi.org/http://dx.doi.org/10.1016/j.ijproman.2010.04.003}{http://dx.doi.org/10.1016/j.ijproman.2010.04.003}.

@@

\hypertarget{ref-Molokken2003}{}
Moløkken, Kjetil, and Magne Jørgensen. 2003. ``A Review of Software
Surveys on Software Effort Estimation.'' Conference Proceedings. In
\emph{Empirical Software Engineering, 2003. Isese 2003. Proceedings.
2003 International Symposium on}, 223--30. IEEE.

@@

\hypertarget{ref-Pinto1988}{}
Pinto, Jeffrey K, and Dennis P Slevin. 1988. ``Critical Success Factors
Across the Project Life Cycle.'' Conference Proceedings. In. Project
Management Institute.

@@

\hypertarget{ref-Shepperd1996}{}
Shepperd, Martin, Chris Schofield, and Barbara Kitchenham. 1996.
``Effort Estimation Using Analogy.'' Conference Proceedings. In
\emph{Proceedings of the 18th International Conference on Software
Engineering}, 170--78. IEEE Computer Society.

@@

\hypertarget{ref-Trost2003}{}
Trost, Steven M, and Garold D Oberlender. 2003. ``Predicting Accuracy of
Early Cost Estimates Using Factor Analysis and Multivariate
Regression.'' Journal Article. \emph{Journal of Construction Engineering
and Management} 129 (2): 198--204.

>>>>

Financial risk is taken on by a consulting company when they offer a
fixed fee to complete a project. In competitive climates, clients are in
the position to request fixed price quotes from several consultants
before a project is awarded. Hence, initial quoted fees must be
competitive but also accurately cover consultant salaries and business
costs.

How do consultants make this accurate yet competitive fee estimate? This
varies from industry to industry, but typically a consulting manager has
experience in the type of project they are quoting and, after reviewing
the project details, can use a combination of intuition, analysis of
project details, and rules of thumb. Unfortunately, some projects priced
using this method can blow out of budget, causing significant financial
losses to the consultant.

Cost blowouts of complex projects has a history, particularly documented
in the It and construction industry. A review of 6 IT project surveys
between 1984 and 1994 by Moløkken \& Jørgensen (2003) revealed that
60-80\% of IT projects encountered effort and/or schedule overruns,
where the average overrun was by 30 to 40\%. A study on large-scale
infrastructure projects over the past seventy years revealed that cost
forecasts consistently underestimated the cost of rail projects by an
average of 44.7\%, the cost of bridge and tunnel projects by 33.8\% and
the cost of road projects by 20.4\% (Flyvbjerg 2007).

@@

This work aims to help managers provide better fixed price quotes by
predicting whether a new project will be profitable or not, based on an
algorithm trained on all previous project data from the same company.
The manager could take multiple actions based on the algorithms output
including not tendering for a project predicted to be loss-making,
providing a high fixed price, or restructuring the contract so that it
is not 100\% fixed price. This research tests the simple case of
rejecting all projects predicted as loss-making.

@@

Project data from a single case study consulting company in the
construction industry was used to build the predictive algorithm, trial
several methods, and test their impact on the company's bottom line.
Twelve years of passively collected data that described each project in
terms of their clients, invoice history, employee hours and technical
details was scraped from the company's internal database. Customer
relationship management (CRM) software provided accesses to the database
to all employees. Employees completed daily timesheets, alotting their
hours to certian projects. Project managers input technical information
and client details for each project, while the finance team tracked
invoicing and additional project costs such as printing and transport in
the same system. On top of this information, the CRM had the hourly cost
of each level of employee. The cleaned dataset consisted of 2364
projects between 2004 and 2012. 70\% of projects were fixed price and
20\% of all projects made a loss after taking into account business
costs.

Once several methods were built, to assess whether the model would
improve the case study company's profitability, a simple managerial
decision framework was applied: if the algorithm predicted the project
to be loss making, the project was considered `rejected' and all profits
or \emph{losses} of rejected projects were discarded. The remaining
projects' profits and losses were summed to give a revised overall
profit over the last 12 years of projects. This was compared to the
summed profits and losses from all projects completed over the last 12
years. The best algorithm rejected enough loss-making projects to
improve overall profits by 9\%.

@@

For the case study company, training the algorithm to predict actual fee
was not possible because the projects were too varied, and there was
inadequate information describing the size and scope of each project in
the database. However, in the literature, similar studies typically did
attempmt to predict project costs of IT consutling projects and final
construction costs of buildings and infrastructure.

@@

Previous studies in the literature have focussed on reporting the
predictive accuracy of their cost estimation models for IT or
construction projects using data from multiple businesses or governments
around the world. In contrast, this research is focussed on determining
how or if models predicting project profitability can improve the bottom
line for a case study business, given data the company has internally
generated.

Reviewing a range of individual studies show that cost estimation models
have generally been built from at most 530 projects (Kim, An, and Kang
2004) (Finnie, Wittig, and Desharnais 1997) - 299 software
development(Pai, McFall, and Subramanian 2013) - 163 software (Shin
2015) - 234 building projects, and often under 100 projects (Finnie,
Wittig, and Desharnais 1997) (Chan and Park 2005) 87 buildings.
Sometimes it is not stated where the project data came from, and some
studies specified the information came from surveys, with responses from
companies and governments around the world. In this study, the dataset
originates from a single company's database, which means the data is
easily accessible to the company and there are 2364 projects, which is
much more than previous studies. A study by Mendes and Kitchenham (2004)
demonstrated that using within company data was also more accurate than
across company data. Their dataset consisted of 67 across company web
projects and 14 within company projects.

Almost all past studies analysed their model performance by predictive
accuracy (reported as R\^{}2 or RMSE) of cost predictions
({\textbf{???}} some stff) and compared these statistics bewteen
methods. However, these results were not translated into a business case
for the impact these models could have on an institution or how the
model could be practically integrated into decision making. For example,
if an institution used the model, how would it be applied to their
decision making process and what magnitude of benefits could they
expect? Saradhi \& Palshikar's (2011) work on employee churn, where
`churn' refers to the number of individuals moving out of a group within
a certain time, presented a good framework for implementing their model.
Theirsupport vector machine (SVM) predictive model identified employees
at high risk of churn. Then a method for determining the value of each
employee in terms of their importance to projects and their monthly
chargeability was described, which enabled high risk employees to be
ranked. Managers then had a clear direction for who to target first to
prevent churn. This extension provided a comprehensive framework for how
business managers could adopt a machine learning algorithm to improve
business operations. It was a valuable addition that is absent from most
cost estimation research, but will be included in this work.

\paragraph{2.2 Methods used in Cost Estimation and Other Business
Applications}\label{methods-used-in-cost-estimation-and-other-business-applications}

Cost estimation literature tended to apply linear regression, neural
networks and sometimes SVM methods, while predictive methods applied to
other business problems investigated a wider range. This study aims to
compare a wider range of methods to past cost estimation studies.

In the construction industry, often only one method was assessed, such
as linear regression, without comparison to other methods (Chan and Park
2005; Elfaki, Alatawi, and Abushandi 2014). Neural Networks started
appearing in literature in the 1990's and Elfaki, Alatawi, and Abushandi
(2014)`s review of cost estimation methods from 2004 to 2014 found that
artificial Neural Networks and SVM's were the most popular machine
learning techniques, but studies tended to only test one method. Some
studies showed that Neural Networks outperformed Linear Regression,
however other studies established they are approximately equal (Kim, An,
and Kang 2004; Attalla and Hegazy 2003). To the authors' knowledge Shin
(2015) documented the only application of ensemble tree methods (Boosted
Trees) to cost estimation in construction projects. They found Boosted
Trees slightly outperformed Neural Networks, but not significanatly.

In the software cost estimation literature, Linear Regression was the
most popular method, however multiple studies showed that Neural
Networks definitively outperform regression models (Finnie, Wittig, and
Desharnais 1997; Pai, McFall, and Subramanian 2013; Matson and
Mellichamp 1993). Despite the higher performance, Neural Networks were
criticised for not providing reasoning or structure behind their
predictions (black-box predictor) (Finnie, Wittig, and Desharnais 1997).

The review of methods applied to the construction and IT industry
highlighted the use of Linear Regression, Neural Networks, SVM's and in
one case Boosted Trees, however research on other business problems
utilised a wider range of methods. In financial credit scoring, a study
by Brown and Mues (2012) found that Random Forests and Boosted Trees
consistently outperformed Neural Networks in classification. In
addition, ensemble trees such as Random Forests and Boosted Trees can
output partial dependency plots and variable importance measures that
provide insight into the model's calculations. This is a valuable asset
oer black box predictors. Kumar and Ravi (2007) performed a detailed
review of statistical and machine learning techniques applied over 37
years in the context of bankruptcy prediction in banks and reported that
although SVM's performed well, they are often complex and slow,
requiring a great deal of memory. They also assessed method blending
techniques, which refers to combinations of two completely different
algorithms, and found they can often outperform individual methods such
as Linear Regression or Boosted Trees alone.

To date, Linear Regression, Neural Networks, and Support Vector Machines
(SVM's) have been comprehensively applied to cost estimation. This
research does not test black-box predictors (Neural Networks and SVM's)
because insight into the model's predictions is crucial for facilitating
user uptake in the case study company. Ensemble tree methods such as
Random Forests and Boosted Trees will be used instead, as they can
perform as well as Neural Networks while providing insight to their
output (Caruana and Niculescu-Mizil 2006). Blending of multiple
predictive models is also applied in this work and has not yet been done
in cost estimation.

@@

In comparison to previous studies, this project advances the body of
work on cost estimation in two ways. First, it focusses on building a
model that is practical and profitable for a business to implement by
using the case study's interenal database as well as testing a
managerial decision framework based off the model output. This framework
presents a clear measure of expected benefits provided by adopting the
profitability model into a business' decision process. Secondly,
advanced predictive methods are trialled which have been minimally
tested on this problem. These include ensemble decision tree methods,
such as Random Forests and Boosted Trees, and blends of multiple models
built from different methods.

@@

\hypertarget{refs}{}
\hypertarget{ref-Attalla2003}{}
Attalla, Mohamed, and Tarek Hegazy. 2003. ``Predicting Cost Deviation in
Reconstruction Projects: Artificial Neural Networks Versus Regression.''
Journal Article. \emph{Journal of Construction Engineering and
Management} 129 (4): 405--11.

@@

@@

@@

@@

@@
